Gradient Descent for Numerical MAX-CSPs

### 2022-05-06 proj/paper/paper.tex
remove homeworks

Many real world problems can be modeled as constraint satisfaction problems (CSPs),
ranging from logic puzzles like sudoku to machine learning. Briefly, a
constraint satisfaction problem is the task of finding a valid assignment of
variables which satisfy a set of constraints. Many classic problems in
theoretical computer science fit this model, such as the 3-SAT and circuit SAT
problems. As such, large amounts of work have been
devoted to studying CSPs and how to solve them, yielding many exact and
approximation algorithms.  In this paper we examine a lesser studied variant of
CSP, called numerical MAX-CSPs. In this variant each variable is a real number
and all of the constraints are linear inequalities, the goal then is to satisfy
as many linear inequalities as possible. We find that while it is less
studied than other CSPs, numerical MAX-CSPs have direct applications to 
a variety real world problems, such as solving over-constrained LP's and 
tuning full text search engines.

@@

Gradient descent is an optimization technique which has gained much attention
in recent years due to  its use in training neural
networks for machine learning tasks such as computer vision and speech
recognition. Despite the focus on machine learning, gradient descent is a
general purpose optimization procedure, capable of optimizing arbitrary
differentiable functions. Many variations have been proposed in recent years,
however each technique uses the same core idea. Given a function $f$ to minimize and
current point $x$, compute the gradient $\nabla(f(x))$
and update $x$ taking a step in a direction of $-\eta \nabla(f(x))$. Repeat
this process until the minimum of the function is found.

If $f(x)$ is convex, then gradient descent is guaranteed to find an optimal
solution (given that $\eta$ is step to an appropriate constant). If $f(x)$ is not convex, the gradient descent will converge to a
local minimum. Although in the non-convex case, gradient descent is not
guaranteed, and frequently won't, find the global minimum, it is still useful for
approximating the global minimum of functions which are cannot be handled by solvers.
This makes it an invaluable tool for optimizing functions which don't admit
exact optimization techniques such as those used for linear and
quadratic programs.

@@

We note that this problem can be is readily formulated as a mixed integer
linear program (MILP) with binary indicator variables. The general problem of
mixed integer linear programming is NP-Hard (by reduction to 0-1
integer programming which is NP-Complete \cite{karp_1972}). While there has
been little work directly addressing numerical MAX-CSPs, there has been a
large body of literature aiming to improve MILP solvers. In particular, there
has been a lot of attention given to improving the branch and bound algorithms
used by many MILP solvers. These improvements typically aim to provide better
branching heuristics, such as strong branching
\cite{strong_branching_paper}. More recent work has examined machine learning
for better branching heuristics \cite{bengio_lodi_prouvost_2021}
\cite{liberto_kadioglu_leo_malitsky_2016}, either to approximate the strong
branching using machine learning (to reduce the computation cost) or to learn
 completely new branching heuristics which will reduce the number of nodes 
 explored to find an optimal solution.

@@

Numerical MAX-CSPs have a wide variety to applications from debugging
infeasible linear programs to machine learning classification, however we are
interested in one particular problem, which is tuning the weights for
full text search engines.  Full text search engines are the backbone of 
many data management and web applications and ergo are a ubiquitous in 
everyday life. These search engines (such as Apache Lucene) are built for
efficient retrieval of top-k documents based on a TF/IDF based scoring metric
which is dot product between two sparse vectors $q^Td =
score$\cite{lucene_history} \cite{WAND_paper} \cite{block_max_WAND_paper}.
While this default scoring gives decent results out of the box, it is
frequently augmented by re-weighting the query $q$, with some weight vector $w$
changing to scoring to be $(q\odot w)^Td = score$ (where $\odot$ is element-wise multiplication).
This allows for boosting of certain terms or fields to improve the quality of
search results while not changing the underlying search algorithm.

@@

$$
f(w) = \langle \text{HardTanh}(Aw - \gamma), \overrightarrow{1}\rangle 
$$

@@

We ran our experiments on a server with an AMD Ryzen Threadripper 2950X
(16c/32t) processor with 64GB of RAM running pop\_os 18.04. Our numerical MAX-CSP
instances are generated from real world datasets, with the number of
constraints ranging from 7.7K to 6.0M.  For MILP we leverage Gurobi, a
commercial solver with an academic license. We choose Gurobi because it is
consistently one of the top performing solvers \cite{benchmark} with an
academic or free license. Both solutions where given a timeout of 2700 seconds,
corresponding to 24 hours of CPU time with 32 threads.

\begin{table}[ht!]
\caption{Dataset statistics}
\begin{tabular}{|l|r|r|r|}
\toprule
Dataset &  \# of Constraints &  \# of Columns &  \% Non-Zero \\
\midrule
         $D_0$ &                   7.7K &                 14 &         66.24 \\
         $D_1$ &                  19.3K &                 18 &         61.46 \\
         $D_2$ &                 111.9K &                 18 &         72.83 \\
         $D_3$ &                 196.2K &                 30 &         51.62 \\
         $D_4$ &                3.7M    &                124 &         25.09 \\
         $D_5$ &                6.0M    &                 22 &         68.58 \\
\bottomrule
\end{tabular}
\end{table}

@@

In this paper we have only taken a small look at the possible optimization
techniques that could be applied. The most obvious variation that could experimented with
is stochastic gradient descent. This variation has the potential for two major
benefits. First, it can decrease both runtime and memory requirements
of the solution further by no longer requiring the entire dataset be processed
on each gradient update. Second, the stochastic version may be able to find
better solutions as it is much more likely that it could escape worse local
minima, instead of relying mostly on having a good starting point.

>>>>

Many real world problems can be modeled as constraint satisfaction problems (CSPs),
ranging from logic puzzles like sudoku to machine learning. Briefly, a
constraint satisfaction problem is the task of finding a valid assignment of
variables which satisfy a set of constraints. Many classic problems in
theoretical computer science fit this model, such as the 3-SAT and circuit SAT
problems. As such, large amounts of work have been
devoted to studying CSPs and how to solve them, yielding many exact and
approximation algorithms \cite{approx_for_csp_paper}. In this paper we examine a lesser studied variant of
CSP, called numerical MAX-CSPs. In this variant each variable is a real number
and all of the constraints are linear inequalities, the goal then is to satisfy
as many linear inequalities as possible. We find that while it is less
studied than other CSPs, numerical MAX-CSPs have direct applications to 
a variety real world problems, such as solving over-constrained LP's and 
tuning full text search engines.

@@

Gradient descent is an optimization technique which has gained much attention
in recent years due to  its use in training neural
networks for machine learning tasks such as computer vision and speech
recognition. Despite the focus on machine learning, gradient descent is a
general purpose optimization procedure capable of optimizing arbitrary
differentiable functions. Many variations have been proposed in recent years,
however each technique uses the same core idea. Given a function $f$ to minimize and
current point $x$, compute the gradient $\nabla(f(x))$
and update $x$ by taking a step in a direction of $-\eta \nabla(f(x))$. Repeat
this process until the minimum of the function is found.

If $f(x)$ is convex, then gradient descent is guaranteed to find the global minimum of the function
(given that $\eta$ is set to an appropriate constant). If $f(x)$ is not convex, then gradient descent will converge to a
local minimum. Although in the non-convex case, gradient descent is not
guaranteed, and frequently won't, find the global minimum, it is still useful for
approximating the global minimum of functions which are cannot be handled by solvers.
This makes it an invaluable tool for optimizing functions which don't admit
exact optimization techniques such as those used for linear and
quadratic programs.

@@

We note that this problem can be is readily formulated as a mixed integer
linear program (MILP) with binary indicator variables. The general problem of
mixed integer linear programming is NP-Hard (by reduction to 0-1
integer programming which is NP-Complete \cite{karp_1972}). While there has
been little work directly addressing numerical MAX-CSPs, there has been a
large body of literature aiming to improve MILP solvers. In particular, there
has been a lot of attention given to improving the branch and bound algorithms
used by many MILP solvers. These improvements typically aim to provide better
branching heuristics, such as strong branching
\cite{strong_branching_paper}. More recent work has examined machine learning
for better branching heuristics \cite{bengio_lodi_prouvost_2021}
\cite{liberto_kadioglu_leo_malitsky_2016}, either to approximate strong
branching using machine learning (to reduce the computation cost) or to learn
 completely new branching heuristics which will reduce the number of nodes 
 explored to find an optimal solution.

@@

Numerical MAX-CSPs have a wide variety to applications from debugging
infeasible linear programs to machine learning classification, however we are
interested in one particular problem, which is tuning the weights for
full text search engines.  Full text search engines are the backbone of 
many data management and web applications and ergo are ubiquitous in 
everyday life. These search engines (such as Apache Lucene) are built for
efficient retrieval of top-k documents based on a TF/IDF based scoring metric
which is dot product between two sparse vectors $q^Td =
score$\cite{lucene_history} \cite{WAND_paper} \cite{block_max_WAND_paper}.
While this default scoring gives decent results out of the box, it is
frequently augmented by re-weighting the query $q$, with some weight vector $w$
changing to scoring to be $(q\odot w)^Td = score$ (where $\odot$ is element-wise multiplication).
This allows for boosting of certain terms or fields to improve the quality of
search results while not changing the underlying search algorithm.

@@

$$
f(w) = \langle \text{HardTanh}(Aw), \overrightarrow{1}\rangle 
$$

@@

We ran our experiments on a server with an AMD Ryzen Threadripper 2950X
(16c/32t) processor with 64GB of RAM running pop\_os 18.04. Our numerical MAX-CSP
instances are generated from real world datasets, with the number of
constraints ranging from 7.7K to 6.0M.  For MILP we leverage Gurobi, a
commercial solver with an academic license. We choose Gurobi because it is
consistently one of the top performing solvers \cite{benchmark} with an
academic or free license. Both solutions where given a timeout of 2700 seconds,
corresponding to 24 hours of CPU time with 32 threads.
\begin{table}[ht!]
	\centering
\caption{Dataset statistics}
\begin{tabular}{|l|r|r|r|}
\toprule
Dataset &  \# of Constraints &  \# of Columns &  \% Non-Zero \\
\midrule
         $D_0$ &                   7.7K &                 14 &         66.24 \\
		 \hline
         $D_1$ &                  19.3K &                 18 &         61.46 \\
		 \hline
         $D_2$ &                 111.9K &                 18 &         72.83 \\
		 \hline
         $D_3$ &                 196.2K &                 30 &         51.62 \\
		 \hline
         $D_4$ &                3.7M    &                124 &         25.09 \\
		 \hline
         $D_5$ &                6.0M    &                 22 &         68.58 \\
\bottomrule
\end{tabular}
\end{table}

@@

In this paper we have only taken a small look at the possible optimization
techniques that could be applied. The most obvious variation that could experimented with
is stochastic gradient descent. This variation has the potential for two major
benefits. First, it can decrease both runtime and memory requirements
of the solution by no longer requiring the entire dataset be processed
on each gradient update. Second, the stochastic version may be able to find
better solutions as it is much more likely that it could escape worse local
minima, instead of relying mostly on having a good starting point.
